Alex's Notes and Text Fragments for the JSS Manuscript

### 2021-08-04 paper/paper.Rmd
revised and reorganized step-by-step section, removed repeated information

We will use a well-known case study to demonstrate the application of spatial resampling techniques.
This binary classification task on landslide susceptibility in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`.

Because the spatial grouping of the $k$-means-based approach (`"spcv_coords"`) (Figure&nbsp;\@ref(fig:vis-spcv)) contrasts visually very well with non-spatial (random) partitioning (Figure&nbsp;\@ref(fig:vis-nspcv)), this method was chosen to showcase the 2D plotting capabilities of \pkg{mlr3spatiotempcv}.

Besides the required arguments of the \pkg{mlr3} superclass `Task`, the `TaskClassifST` / `TaskRegrST` classes from package \pkg{mlr3spatiotempcv}  also require the following arguments during construction which must be passed via a named list `extra_args`:

- `coordinate_names`: Names of the features that represent the spatial coordinates.
  This is automatically inferred when a `sf` object is passed.
- `coords_as_features`: Whether the coordinates should be used as features during modeling.
  By default coordinates are not used as predictors.
- `crs`: The coordinate reference system of the data. Valid inputs are PROJ strings or EPSG codes in the format `ESPG:<code>`.

At first all necessary R packages are loaded and a lower verbosity is set to keep the output tidy.
Also a seed is set for deterministic execution and the \pkg{mlr3spatiotempcv} built-in task `"ecuador"` is loaded via `tsk()`.

@@

To set up the task `"ecuador"` manually, we would have to proceed as follows:

```{r create-task}
# equivalent to 'tsk("ecuador")':
b = mlr3::as_data_backend(ecuador)
b$hash = "_mlr3_tasks_ecuador_" # optional
task = TaskClassifST$new(
  id = "ecuador", b, target = "slides", positive = "TRUE",
  extra_args = list(
    coordinate_names = c("x", "y"), coords_as_features = FALSE,
    crs = "EPSG:32717")
)
```

To show the effect of spatial autocorrelation on performance estimates in cross-validation, we will apply spatial and non-spatial cross-validation to the prediction of landslides using the random forest technique (`"classif.ranger"`).
Specifically, we use spatial partitioning based on $k$-means clustering of the coordinates (method `"repeated_spcv_coords"`) and compare its results to a non-spatial (random) cross-validation (`"repeated_cv"`), both with (only) four folds and two repetitions for demonstration.
We choose the area under the ROC curve (`"classif.auc"`) as the performance measure for this classification task.

@@

Next, the random forest learner (`lrn("classif.ranger")`) is initialized with default hyperparameters and the prediction type is set to `"probability"` due to the binary response variable.
A set of commonly used learners is available in package \pkg{mlr3learners} [@mlr3learners], including the random forest implementation of @ranger.

@@

A four-fold, two times repeated non-spatial resampling object is constructed and instantiated via method `$instantiate()`.
Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the task and learner objects created earlier to execute the model assessment.

The results from the CV are then aggregated using the measure `"classif.auc"`, which is the \pkg{mlr3} package notation for the AUROC measure.

@@

The same approach as for non-spatial resampling is used for spatial resampling, expect replacing method `"repeated_cv"` with `"repeated_spcv_coords"`.

@@

```{r vis-spcv, fig.cap="Spatial leave-one-block-out partitioning using coordinate-based clustering to create roughly equally sized polygonal blocks. Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8}
resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)

autoplot(resampling_sp, task, fold_id = c(1:2), size = 0.8) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

```{r vis-nspcv, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8}
resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)

autoplot(resampling_nsp, task, fold_id = c(1:2), size = 0.8) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

The non-spatial CV estimate of AUC (0.77) is around 0.15 better compared to the spatial CV estimate of 0.61.
The better performance results from over-optimism in the estimation with non-spatial CV.
Spatial CV, in contrast, provides a better measure of a model's ability to generalize from the training sample --- in the Ecuador case study, from the hillslopes and historical landslides in the training sample.
It is also expected that spatial CV results better reflect the performance of a model on new, spatially separate data.
The magnitude of this difference may depend on the dataset, the strength on spatial or spatiotemporal autocorrelation, and the learner itself.
Algorithms with a higher tendency to overfit to the training set will tend to have a larger spread in such scenarios.

>>>>

We use a well-known case study to demonstrate the application of spatial and non-spatial resampling techniques for model assessment in \pkg{mlr3spatiotempcv}.
The objective of landslide susceptibility modeling is to predict how prone to landslide initiation a location is.
Models are fitted to historical landslide occurrences, but they need to learn generalizable relationships between predisposing variables and the response as opposed to perfectly reproducing or memorizing the historical distribution. 
This binary classification task on landslides in Ecuador [@muenchow2012] is available as a built-in task via `tsk("ecuador")`, but we generate it from the learning sample.
Random forest is used as a classifier, and the area under the ROC curve (AUROC) as the performance measure.

Spatial CV is implemented in the form of leave-one-block-out CV using coordinate-based $k$-means clustering to generate  irregularly shaped blocks of roughly equal size. 
This approach is better suited for the shape of the present study area than a rectangular partitioning.
Figures&nbsp;\@ref(fig:vis-spcv)) and&nbsp;\@ref(fig:vis-nspcv)) show the contrasting distributions of training and test samples.
For demonstration purposes we use only four CV folds and two repetitions.

## Task preparation

In \pkg{mlr3}, machine-learning tasks with their respective dataset and response variable are represented by objects of the class `Task`. \pkg{mlr3spatiotempcv}'s spatial and spatiotemporal machine-learning tasks are derived from this superclass. 
Specifically, the `TaskClassifST` and `TaskRegrST` classes for classification and regression tasks require several additional arguments that must be passed as a named list using the `extra_args` argument:

- `coordinate_names`: Names of the features that represent the spatial coordinates.
  This is automatically inferred when a `sf` object is passed.
- `coords_as_features`: Whether the coordinates should be used as features; by default they are not.
- `crs`: The coordinate reference system of the data as a PROJ string or EPSG code in the format `ESPG:<code>`.

At first all necessary R packages are loaded and a lower verbosity is set to keep the output tidy.
A random-number seed is set for reproducibility.

@@

The task `"ecuador"` is available in \pkg{mlr3spatiotempcv} through `tsk("ecuador")`.
To set it up manually, we proceed as follows:

<!-- AB: TODO: The comment says that b$hash is optional, but the reader will not know what it is good for. A more verbose comment would be helpful, maybe 'optional task identifier' or whatever the hash name may be good for. -->
```{r create-task}
b = mlr3::as_data_backend(ecuador)
b$hash = "_mlr3_tasks_ecuador_" # optional
task = TaskClassifST$new(
  id = "ecuador", b, target = "slides", positive = "TRUE",
  extra_args = list(
    coordinate_names = c("x", "y"), coords_as_features = FALSE,
    crs = "EPSG:32717")
)
```

@@

Next, the random forest learner (`"classif.ranger"`) is initialized with default hyperparameters and the prediction type is set to `"probability"` because the model is used for soft classification.
A set of commonly used learners is available in package \pkg{mlr3learners} [@mlr3learners], including the random forest implementation of @ranger.

@@

A resampling is generated in \pkg{mlr3} with the `rsmp()` function, which is used to generate a resampling object for four-fold non-spatial CV with two repetitions.
The object is constructed and instantiated with its `$instantiate()` method.

Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the task and learner objects created earlier to execute the model assessment.
This is the actual, potentially time-consuming CV estimation. With the present settings, eight random forest classifiers are fitted and evaluated in this step.

Model performances are calculated from the CV predictions using the AUROC (`"classif.auc"` in \pkg{mlr3} notation).

@@

The model assessment is now repeated using spatial CV resampling, for which we only need to replace `"repeated_cv"` with `"repeated_spcv_coords"`.

@@

```{r vis-spcv, fig.cap="Spatial leave-one-block-out partitioning using coordinate-based clustering to create roughly equally sized polygonal blocks. Due to space limitations only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8}
autoplot(resampling_sp, task, fold_id = c(1:2), size = 0.8) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

```{r vis-nspcv, fig.cap="Random (non-spatial) four-fold CV partitioning. Only the first two folds of the first repetition are shown.", fig.pos="ht", fig.height=6, fig.width=8}
autoplot(resampling_nsp, task, fold_id = c(1:2), size = 0.8) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

The non-spatial CV estimate of AUC (0.76) is around 0.13 better compared to the spatial CV estimate of 0.64.
The better performance results from over-optimism in the estimation with non-spatial CV.
Spatial CV, in contrast, provides a better measure of a model's ability to generalize from the training sample --- in this case study, from the specific hillslopes and historical landslides in the training sample.
It is also expected that spatial CV results better reflect the performance of a model on new, spatially separate data.
The magnitude of the difference between spatial and non-spatial CV estimates may depend on the dataset, the strength on spatial or spatiotemporal autocorrelation, and the learner itself.
Algorithms with a higher tendency to overfit to the training set will tend to have a larger spread in such scenarios.
